\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2019-02-21 archive.tex
updated learning curves

An old dream within AI is a machine that learns and reasons by
writing its own programs.  This vision stretches back to the
1960's~\cite{solomonoff1964formal} and and is central to much of
what it would take to build machines that learn and think like humans.
Computational models of cognition often explain the flexibility and
richness of human thinking in terms of program learning: from everyday
thinking and problem solving (motor program induction as an account of
recognition and generation of handwriting and
speech~\cite{lake2015human}; functional programs as a model of natural
language semantics~\cite{SOMETHING}) to learning problems that unfold
over longer developmental time scales: the child's acquisition of
intuitive theories (of kinship, taxonomy, etc.)~\cite{Ullman2012} and
natural language grammar~\cite{DBLP:journals/cogsr/SchmidK11}, to name
just a few.  An outstanding challenge, however, is to engineer
program-learners that display the same level of domain-generality as
the humans they are meant to model.

Recent program-learning systems developed within the AI and machine
learning community are impressive along many dimensions, authoring
programs for problem domains like drawing
pictures~\cite{spiral,ellis2017learning}, transforming
text~\cite{gulwani2011automating} and numerical
sequences~\cite{balog2016deepcoder}, and reasoning over common sense
knowledge bases~\cite{muggleton2015meta}.  These systems work in
different ways, but typically hinge upon a carefully hand-engineered
Domain Specific Language (DSL).  The DSL restricts the space of
programs to contain the kinds of concepts needed for one specific
domain.  For example, a picture-drawing DSL could include concepts
like circles and spirals, and a DSL for numerical sequences could
include sorting and reversing lists of numbers.  Modern systems also
learn how to efficiently deploy the DSL on new
problems~\cite{devlin2017robustfill,balog2016deepcoder,NGDS}, but --
unlike human learners -- do not discover the underlying system of
concepts needed to navigate the domain.

We contribute a program-induction system that learns the
domain-specific concepts (DSL) while jointly learning how to use those
concepts.  This joint learning problem models two complementary
notions of domain expertise: (1) domain experts have at their disposal
a powerful, yet specialized repertoire of concepts and abstractions
(analogous to the DSL) while also (2) having accurate intuitions about
when and how to use those concepts to solve new problems.
Representative domains,
along with DSLs we learn for them,
are shown in Figure~\ref{exampleDSL}.

We call our system `DreamCoder' because it acquires these two kinds of
expertise through a novel kind of wake/sleep or `dream'
learning~\cite{hinton1995wake}, iterating through a wake cycle --
where it solves problems by writing programs -- and a pair of sleep
cycles, both of which are loosely biologically inspired by actual
sleep.  The first sleep cycle, which we refer to as
\textbf{consolidation}, grows the DSL by replying experiences from
waking and consolidating them into new code abstractions.  This cycle
is inspired by the formation of abstractions during sleep memory
consolidation~\cite{DUDAI201520}.  The second sleep cycle, which we
refer to as \textbf{dreaming}, improves the agents knowledge of how to
write programs by training a neural network to help search for
programs. The neural net is trained on replayed experiences as well as
`fantasies', or samples, from the DSL.  These two kinds of dreams are
inspired by the distinct episodic replay and hallucination components
of dream sleep~\cite{fosse2003dreaming}.

@@

    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/list_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/list_time.png}
    \end{tabular}&

    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/tower_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/tower_time.png}
    \end{tabular}
  \end{tabular}
  \caption{Test set performance across wake/sleep iterations. Each curve is a run with a different random seed. Teal: Full model. Orange: Dreaming only (no DSL learning). Purple: Consolidation only (no recognition model). Search time plots show solid lines (time averaged over all tasks) and dotted lines (time averaged over solved tasks).}\label{learningCurves}      
\end{figure}

@@

>>>>

An old dream within AI is a machine that learns and reasons by writing
its own programs.  This vision stretches back to the
1960's~\cite{solomonoff1964formal} and, if fully realized, could bring
us much closer to machines that learn and think like humans.
Computational models of cognition often explain the flexibility and
richness of human thinking in terms of program learning: from everyday
thinking and problem solving (motor program induction as an account of
recognition and generation of handwriting and
speech~\cite{lake2015human}; functional programs as a model of natural
language semantics~\cite{SOMETHING}) to learning problems that unfold
over longer developmental time scales: the child's acquisition of
intuitive theories (of kinship, taxonomy, etc.)~\cite{Ullman2012} and
natural language grammar~\cite{DBLP:journals/cogsr/SchmidK11}, to name
just a few.  An outstanding challenge, however, is to engineer
program-learners that display the same level of domain-generality as
the humans they are meant to model.

Recent program-learning systems developed within the AI and machine
learning community are impressive along many dimensions, authoring
programs for problem domains like drawing
pictures~\cite{spiral,ellis2017learning}, transforming
text~\cite{gulwani2011automating} and numerical
sequences~\cite{balog2016deepcoder}, robot motion
planning~\cite{devlin2017neural}, and reasoning over common sense
knowledge bases~\cite{muggleton2015meta}.  These systems work in
different ways, but typically hinge upon a carefully hand-engineered
Domain Specific Language (DSL).  The DSL restricts the space of
programs to contain the kinds of concepts needed for one specific
domain.  For example, a picture-drawing DSL could include concepts
like circles and spirals, and a DSL for numerical sequences could
include sorting and reversing lists of numbers.  Modern systems also
learn how to efficiently deploy the DSL on new
problems~\cite{devlin2017robustfill,balog2016deepcoder,NGDS}, but --
unlike human learners -- do not discover the underlying system of
concepts needed to navigate the domain.

We contribute a program-induction system that learns the
domain-specific concepts (DSL) while jointly learning how to use those
concepts.  This joint learning problem models two complementary
notions of domain expertise: domain experts have at their disposal
a powerful, yet specialized repertoire of concepts and abstractions
(analogous to the DSL) while also having accurate intuitions about
when and how to use those concepts to solve new problems.
Representative domains,
along with DSLs we learn for them,
are shown in Figure~\ref{exampleDSL}.

We call our system `DreamCoder' because it acquires these two kinds of
expertise through a novel kind of wake/sleep or `dream'
learning~\cite{hinton1995wake}, iterating through a wake cycle --
where it solves problems by writing programs -- and a pair of sleep
cycles, both of which are loosely biologically inspired by actual
sleep.  The first sleep cycle, which we refer to as
\textbf{consolidation}, grows the DSL by replying experiences from
waking and consolidating them into new code abstractions.  This cycle
is inspired by the formation of abstractions during sleep memory
consolidation~\cite{DUDAI201520}.  The second sleep cycle, which we
refer to as \textbf{dreaming}, improves the agents knowledge of how to
write code by training a neural network to help search for
programs. The neural net is trained on replayed experiences as well as
`fantasies', or samples, from the DSL.  These two kinds of dreams are
inspired by the distinct episodic replay and hallucination components
of dream sleep~\cite{fosse2003dreaming}.

@@

    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/list_hard_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/list_hard_time.png}
    \end{tabular}&

    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/tower_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/tower_time.png}
    \end{tabular}&

    \includegraphics[width = 5cm]{figures/learningCurves/regular}
  \end{tabular}
  \caption{Test set performance across wake/sleep iterations. Each curve is a run with a different random seed. Teal: Full model. Orange: Dreaming only (no DSL learning). Purple: Consolidation only (no recognition model). Search time plots show solid lines (time averaged over all tasks) and dotted lines (time averaged over solved tasks).}\label{learningCurves}      
\end{figure}

@@

\subsection{\system and the Exploration-Compression family of algorithms}

Our work sits within the Exploration-Compression (EC) family of
algorithms.  EC~\cite{Dechter:2013:BLV:2540128.2540316} is a program
induction framework where an agent alternates between searching, or
`exploring', the space of programs, and then updating its search
procedure by compressing programs found during exploration. \system
grew directly out of research on EC-style systems and both of our
sleep phases can be interpreted as a kind of compression:
consolidation aims to compactly refactor code, while the recognition
model aims to encode a program in as few bits as possible, conditioned
on a task. Other offshoots of EC include our previous work,
EC$^2$~\cite{ecc}, the direct predecessor of \systemEnding; and the
neurosymbolic framework in~\cite{lazaro2019beyond}. Closely aligned ideas
go much further back~\cite{solomonoff1989system,schmidhuber2004optimal}.
